---
title: "Report on Motorcycle Accident Deaths in Colombia"
author: "Olugbenga Asaolu" # Remember to add your name!
output:
  prettydoc::html_pretty:
    theme: architect
date: "`r Sys.Date()`"
---

::: {.hidden}

```{r include = FALSE}
# Set chunk options
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r}
# TASK 1: LOAD THE REQUIRED PACKAGES
if(!require(pacman)) install.packages("pacman")
pacman::p_load(tidyverse, here, janitor, reactable)
```

```{r}
# TASK 2: IMPORT THE DATA
col_accidents <- read_csv(here("data", "motorcycle_accidents_colombia.csv"))
```

```{r}
# TASK 3: CLEAN THE COLUMN NAMES 
col_accidents_clean <- col_accidents %>% 
  clean_names()
```

```{r}
# TASK 4: RENAME AND TRANSFORM VARIABLES
col_accidents_clean2 <- col_accidents_clean %>%
  rename(position = condition) %>%
  mutate(gender = str_to_title(gender))
```

```{r}
# TASK 5: CREATE A SUBSET
col_accidents_subset <- col_accidents_clean2 %>%
  select(id, gender, age, age_group, position) %>%
  filter(position != "unknown")
```

```{r}
# TASK 6: SUMMARIZE DATA
age_sex_count <- col_accidents_subset %>%
  count(age_group, gender)
```

```{r}
# TASK 7: VISUALIZE SUMMARY DATA
death_plot2 <- ggplot(age_sex_count, aes(x = age_group, y = n, fill = gender)) +
  geom_col(position = "stack") +
  geom_text(aes(label = n), 
            position = position_stack(vjust = 0.5), 
            color = "white", size = 3) +
  labs(
    title = "Number of Motorcycle Accident Deaths by Age Group and Gender",
    subtitle = "Stacked bar chart showing the distribution of fatalities by demographic group",
    caption = "Data source: Motorcycle accident fatalities dataset, Colombia",
    x = "Age Group",
    y = "Number of Deaths",
    fill = "Gender"
  ) +
  theme_minimal()

death_plot2
```

```{r}
# TASK 8: EXPORT YOUR OUTPUTS
write_csv(age_sex_count, here("outputs", "age_sex_count.csv"))

ggsave(
  filename = here("outputs", "death_plot.png"),
  plot = death_plot2,
  width = 8,
  height = 6,
  dpi = 300
)
```


:::

# Introduction

The dataset analysed in this report provides information about **`r nrow(col_accidents_subset)`** deaths caused by motorcycle accidents in Medellín, a Colombian city, from 2012 to 2015. The data is sourced from medical and police certificates and can be accessed [here](https://zenodo.org/record/4836304).

# Age-sex distribution of motorcyle deaths

In this analysis, we looked at the number of deaths per age and sex group.

```{r}
# Print your graph here.
death_plot2
```

The age group with the highest number of motorcycle accident deaths among males is `r age_sex_count %>% filter(gender == "Male") %>% arrange(desc(n)) %>% slice(1) %>% pull(age_group)` with `r age_sex_count %>% filter(gender == "Male") %>% arrange(desc(n)) %>% slice(1) %>% pull(n)` deaths.


```{r}
# TASK 9: ADD INLINE CODE AND COMMENTARY 

## Each of the 3 statistics in this code chunk will be quoted in the text below.

max_sex_group <- age_sex_count %>% 
  filter(n == max(n)) %>% 
  pull(gender)

max_age_group <- age_sex_count %>% 
 filter(n == max(n)) %>% 
  pull(age_group)

max_n_deaths <- age_sex_count %>% 
  summarise(max_deaths = max(n)) %>% 
  pull(max_deaths)
```

The age group with the most fatal accidents was `r paste0(max_sex_group, "s")` aged `r max_age_group`, with `r max_n_deaths` deaths.

## Analysis of Driver Position 

We further analyzed the position of the driver involved in these accidents for the age-sex group with the most fatal accidents. 

The table below shows the breakdown of accidents by driver position:

```{r}
# TASK 10: SUMMARIZE DIRVER POSITION DATA 
# Filter and summarize
driver_position_subset <- col_accidents_subset %>%
  filter(gender == "Male", age_group == "20-24") %>%
  count(position, name = "number_of_deaths") %>%
  arrange(desc(number_of_deaths))

# Render summary table using reactable
reactable(driver_position_subset,
          columns = list(
            position = colDef(name = "Driving Position"),
            number_of_deaths = colDef(name = "Number of Deaths")
          ),
          bordered = TRUE,
          striped = TRUE,
          highlight = TRUE,
          defaultPageSize = 5,
          showPageSizeOptions = TRUE)
```
# Time Series Analysis of Yearly deaths

In this analysis, we looked at the yearly number of deaths .

```{r}
# Extract year from the date_of_death column and summarize
yearly_deaths <- col_accidents_clean2 %>%
  filter(!is.na(date_of_death)) %>%
  mutate(year = year(date_of_death)) %>%
  count(year, name = "number_of_deaths") %>%
  arrange(year)

# Create the plot
yearly_death_plot <- ggplot(yearly_deaths, aes(x = factor(year), y = number_of_deaths)) +
  geom_col(fill = "#20B2AA") +
  geom_text(aes(label = number_of_deaths), vjust = -0.3, size = 3.5) +
  labs(
    title = "Motorcycle Accident Deaths in Colombia by Year",
    subtitle = "Time series showing total number of deaths each year",
    caption = "Data source: Motorcycle accident fatalities dataset (2012–2015)",
    x = "Year",
    y = "Number of Deaths"
  ) +
  theme_minimal()

yearly_death_plot
```

